---
title: "repdata_assignment1.rmd"
author: "Mogens Yde-Andersen"
date: "June 11, 2015"
output: 
  html_document:
    toc: true
    theme: united
  md_document:
    toc: true
---

###Loading and preprocessing the data

1.Load the data (i.e. read.csv())

```{r, echo=TRUE, cache=TRUE}
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", "activity.zip")
unzip("activity.zip")
rawdata <- read.csv("activity.csv", header=TRUE, sep = ",", dec = ".")
```

###Process/transform the data (if necessary) into a format suitable for your analysis

I did not process anything at this stage.

What is mean total number of steps taken per day?

1.Calculate the total number of steps taken per day

```{r}
sumstepsperday <- tapply(rawdata$steps, rawdata$date, FUN=sum, na.rm=TRUE)
head(sumstepsperday)
```

2.Make a histogram of the total number of steps taken each day

```{r}
hist(sumstepsperday, breaks=16, col = "lightgreen", border = "darkgreen", ylim=c(0,15), xlim=c(0,25000), xlab="Total number of steps taken each day", main="Histogram: Summarised number of steps taken each day")
```

3.Calculate and report the mean and median of the total number of steps taken per day

```{r}
mean(sumstepsperday, na.rm=TRUE)
median(sumstepsperday, na.rm=TRUE)
```

The mean of the total number of steps taken each day is 9,354.23, and median of the total number of steps taken each day is 10,395.

###What is the average daily activity pattern?
1.Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
meanstepsperinterval <- aggregate(x=list(steps=rawdata$steps), by=list(interval=rawdata$interval),
                      FUN=mean, na.rm=TRUE)
plot(meanstepsperinterval, type="l", col = "darkgreen",ylim=c(0,225), xlim=c(0,2400), xlab="5 minute intervals located at the time (HHMM) of the day", ylab="Average steps taken per interval", main = "Average steps taken in each 5 minute interval across all days")
```

2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
meanstepsperinterval[which.max(meanstepsperinterval$steps),]
```
The 5 minute interval at 08:35 in the morning is the 5 minute interval across all days, where the most steps on average were taken, 206.1698 steps.

###Imputing missing values

1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
nas <- is.na(rawdata$steps)
table(nas)
```

2.Devise a strategy for filling in all of the missing values in the dataset.

I replace missing values with the mean value of each 5 minute interval across all days.

3.Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
process.value <- function(steps, interval) {
    processed <- NA
    if (!is.na(steps))
        processed <- c(steps)
    else
        processed <- (meanstepsperinterval[meanstepsperinterval$interval==interval, "steps"])
    return(processed)
}
processed.data <- rawdata
processed.data$steps <- mapply(process.value, processed.data$steps, processed.data$interval)
head(processed.data)
```

4.Make a histogram of the total number of steps taken each day

```{r}
sumstepsperdayprocessed <- tapply(processed.data$steps, processed.data$date, FUN=sum, na.rm=TRUE)
head(sumstepsperdayprocessed)
```

```{r}
hist(sumstepsperdayprocessed, breaks=16, col = "lightgreen", border = "darkgreen", ylim=c(0,20), xlim=c(0,25000), xlab="Total number of steps taken each day", main="Histogram: Summarised number of steps taken each day (without NA's)")

```

3.Calculate and report the mean and median of the total number of steps taken per day

```{r}
mean(sumstepsperdayprocessed, na.rm=TRUE)
median(sumstepsperdayprocessed, na.rm=TRUE)
```

The "new" mean of the total number of steps taken each day is 10,766.19, and the "new" median of the total number of steps taken each day is 10,766.19.

Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Yes, the mean og the median of the processed data deviates from the mean and the median of the raw data. In the raw data the NA's are set til the value 0, where as my strategi sets the value af the NA's to mean value of each 5 minute interval across all days. This "moves" all NA observations from the very left in a histogram rightwards and increases both mean and median.

###Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

1.Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

(Note that "Lørdag" and Søndag" are the danish words for "Saturday" and"Sunday" in english.)

```{r}
processed.data$weekday <- weekdays(as.Date(processed.data$date, format = "%Y-%m-%d"))
processed.data$weekday.type <- factor(ifelse(processed.data$weekday=="Lørdag"|processed.data$weekday=="Søndag", "weekend", "weekday"), levels = c("weekday", "weekend"))
processed.data$interval <- as.integer(processed.data$interval)
str(processed.data)
```

2.Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
weekdaysteps <- processed.data[processed.data$weekday.type == "weekday", ]
str(weekdaysteps)
weekendsteps <- processed.data[processed.data$weekday.type == "weekend", ]
str(weekendsteps)
meanweekdaysteps <- aggregate(x=list(steps=weekdaysteps$steps), by=list(interval=weekdaysteps$interval),
                      FUN=mean, na.rm=TRUE)
meanweekendsteps <- aggregate(x=list(steps=weekendsteps$steps), by=list(interval=weekendsteps$interval),
                      FUN=mean, na.rm=TRUE)
par(mfrow = c(2,1), mar = c(2.7,4,1.2,1))
plot(meanweekdaysteps, type="l", col = "darkgreen", ylim=c(0,250), ylab="")
legend("topright", col=("darkgreen"), lty=c(1),  bty="n", c("Weekdays"))
plot(meanweekendsteps, type="l", col = "darkgreen", ylim=c(0,250), ylab="")
legend("topright", col=("darkgreen"), lty=c(1),  bty="n", c("Weekenddays"))
par(mfrow = c(1,1), mar = c(1,2,1,0))
mtext(text="5 minute intervals located at the time (HHMM) of the day", side=1)
mtext(text="Average steps taken per interval", side=2)
mtext(text="Average steps taken in each 5 minute interval across Weekdays and Weekenddays", side=3)
```

Yes, there are differences in activity patterns between weekdays and weekend days. People tend to be more active in mornings on weekdays, where as they are more active around midday and in the afternoons on weekend days. Just before 20:00 (8:00 PM) people are more active on weekdays, where as they are slightly more active just after 20:00 (8:00 PM) on weekend days.